---
title: "TransportationFinal"
author: "Jack Chen"
date: "2024-12-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidycensus)
library(dplyr)
library(cancensus)
library(ggplot2)
library(RColorBrewer)
library(tidyr)
library(knitr)
library(tidyverse)
library(kableExtra)
library(terra)
library(sf)
library(biscale)
library(cowplot)
```
Vector: short variable code
Type: variables are provided as aggregates of female responses, male responses, or total (male+female) responses
Label: detailed variable name
Units: provides information about whether the variable represents a count integer, a ratio, a percentage, or a currency figure
Parent_vector: shows the immediate hierarchical parent category for that variable, where appropriate
Aggregation: indicates how the variable should be aggregated with others, whether it is additive or if it is an average of another variable
Description: a rough description of a variable based on its hierarchical structure. This is constructed by cancensus by recursively traversing the labels for every variable’s hierarchy, and facilitates searching for specific variables using key terms.     

```{r eval=FALSE}
set_cancensus_api_key('CensusMapper_54c41158364bb71fabdde2b3853b84b8', install = TRUE)
```

```{r}
vancensus = get_census(dataset='CA21', regions=list(CMA="59933"),
                       vectors=c("v_CA21_4875","v_CA21_906","v_CA21_251", "v_CA21_11", "v_CA21_74", "v_CA21_1"), #minority, hh inc, elderly population total, 0-14, 15, totPop
                       level='CT', use_cache = FALSE, geo_format = 'sf', quiet = TRUE)
  
```

```{r fig.width = 10, fig.height = 8}
vancensus.f = vancensus %>%
  filter(CSD_UID %in% c(5915020, 5915022, 5915803)) %>%
  filter(`Region Name` != '0251.02') %>%
  rename("totMinority" = "v_CA21_4875: Total visible minority population",
         "medinc" = "v_CA21_906: Median total income of household in 2020 ($)",
         "age65" = "v_CA21_251: 65 years and over",
         "age14" = "v_CA21_11: 0 to 14 years",
         "age15" = "v_CA21_74: 15",
         "totPop21" = "v_CA21_1: Population, 2021") %>%
  mutate("disadvPop" = age65 + age14 + age15,
         "disadvPopPct" = disadvPop/totPop21,
         "minorityPct" = totMinority/totPop21)

vanstations = st_read("./data&code/vanstops.geojson")

ggplot() + 
  geom_sf(data = vancensus.f) +
  geom_sf(data = vanstations) +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) 
```

```{r}
intersection200 = st_read("./data&code/200mIntersection.geojson")
head(as.data.frame(intersection200))

vancensus200 = vancensus.f %>%
  left_join(as.data.frame(intersection200) %>%
              select(name, percentagecover, GeoUID, area, intersectarea), by = "name") %>%
  mutate("covPop" = intersectarea/totPop21,
         "covPctPop" = percentagecover/totPop21) %>%
  mutate("weightedCover" = percentagecover*(totPop21/(sum(vancensus.f$totPop21))))
```

```{r eval=FALSE}
st_write(vancensus200, "./data&code/vancensus.geojson")
```

```{r}
ggplot(vancensus200) +
  geom_sf(aes(fill = weightedCover)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())
```

```{r fig.width = 10, fig.height=7}
minoritycover <- bi_class(vancensus200, x = weightedCover, y = minorityPct, style = "quantile", dim = 3)

minoritymap <- ggplot() +
  geom_sf(data = minoritycover, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "GrPink", dim = 3) +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) +
  bi_theme()

minoritylegend = bi_legend(pal = "GrPink",
                    dim = 3,
                    xlab = "Higher % Station Cover ",
                    ylab = "Higher % Minority ",
                    size = 6)

minorityPlot <- ggdraw() +
  draw_plot(minoritymap, 0, 0, 1, 1) +
  draw_plot(minoritylegend, 0, 0.15, 0.2, 0.2)

minorityPlot
```

```{r fig.width = 10, fig.height=7}
disadvcover <- bi_class(vancensus200, x = weightedCover, y = disadvPopPct, style = "quantile", dim = 3)

disadvmap <- ggplot() +
  geom_sf(data = disadvcover, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "GrPink", dim = 3) +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) +
  bi_theme()

disadvlegend = bi_legend(pal = "GrPink",
                    dim = 3,
                    xlab = "Higher % Station Cover",
                    ylab = "Higher % Disadvantaged",
                    size = 6)

disadvPlot <- ggdraw() +
  draw_plot(disadvmap, 0, 0, 1, 1) +
  draw_plot(disadvlegend, 0, 0.15, 0.2, 0.2)

disadvPlot
```

```{r fig.width = 10, fig.height=7}
inccover <- bi_class(vancensus200, x = weightedCover, y = medinc, style = "quantile", dim = 3)

incmap <- ggplot() +
  geom_sf(data = inccover, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "GrPink", dim = 3) +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) +
  bi_theme()

inclegend = bi_legend(pal = "GrPink",
                    dim = 3,
                    xlab = "Higher % Station Cover",
                    ylab = "Higher Median Income",
                    size = 6)

incPlot <- ggdraw() +
  draw_plot(incmap, 0, 0, 1, 1) +
  draw_plot(inclegend, 0, 0.15, 0.2, 0.2)

incPlot
```

```{r}
stop_times = read.csv("./data&code/translinkdata/stop_times.csv")

vanstations$stop_id = as.numeric(vanstations$stop_id)

head(as.data.frame(vanstations))

van_stop_times <- stop_times %>%
  semi_join(vanstations, by = "stop_id")  

van_stop_times <- van_stop_times %>%
  left_join(vanstations %>% select(stop_id, stop_code, stop_name, stop_lat, stop_lon), by = "stop_id")
```

```{r}
trips = read.csv("./data&code/translinkdata/trips.csv")

trips_filtered <- trips %>%
  semi_join(van_stop_times, by = "trip_id")  # Keeps only trip_ids in stop_times_with_info

van_stop_trips <- van_stop_times %>%
  left_join(trips_filtered %>% select(route_id, service_id, trip_id, trip_headsign, direction_id, block_id, shape_id),
            by = "trip_id")

van_stop_trips = van_stop_trips %>%
  dplyr::select(-stop_headsign, -pickup_type, -drop_off_type, -shape_dist_traveled, -timepoint)
```

```{r}
calendar = read.csv("./data&code/translinkdata/calendar.csv")

van_cal <- van_stop_trips %>%
  left_join(calendar %>% select(service_id, monday, tuesday, wednesday, thursday, friday, saturday, sunday),
            by = "service_id") %>%
  mutate(weekday_avail = monday+tuesday+wednesday+thursday+friday+saturday+sunday)
```

```{r}
dates = read.csv("./data&code/translinkdata/calendar_dates.csv")

# Step 1: Create a count of exceptions for service_id in the dates dataframe
exception_counts <- dates %>%
  group_by(service_id) %>%
  summarize(total_exceptions = n(), .groups = "drop")  %>%
  filter(!(service_id %in% c(1, 2, 3)))

# Step 2: Add operating_dates to final_df_with_calendar
van_translink_trips <- van_cal %>%
  # Left join with exception counts
  left_join(exception_counts, by = "service_id") %>%
  # Create the operating_dates column based on conditions
  mutate(
    operating_dates = case_when(
      !is.na(total_exceptions) ~ total_exceptions, # Use total_exceptions if service_id is in dates and not "1"
      is.na(total_exceptions) & service_id == "1" ~ 84,               # Default to 84 if service_id is "1" and not in dates
      is.na(total_exceptions) & service_id %in% c("2", "3") ~ 18,     # Default to 18 if service_id is "2" or "3" and not in dates
      TRUE ~ 0                                                       # Otherwise, default to 0
    ),
    pctOperating = operating_dates/126*100
  ) %>%
  # Drop the total_exceptions column (optional, if not needed further)
  select(-total_exceptions)
```

```{r}
unique_trips = van_translink_trips %>%
  dplyr::select(-arrival_time, -trip_id, -departure_time, -block_id, -monday, -tuesday, -wednesday, -thursday, -friday, -saturday, -sunday, -shape_id, -stop_sequence) %>%
  unique() #%>%
  #dplyr::select(-service_id)

unique_trip_sums <- unique_trips %>%
  group_by(trip_headsign, stop_id, stop_name) %>%
  summarise(
    weekday_avail_sum = sum(weekday_avail, na.rm = TRUE),
    operating_dates_sum = sum(operating_dates, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(operating_dates_sum = case_when(operating_dates_sum > 126 ~ 126,
                                        TRUE ~ operating_dates_sum)) 
# Upon inspection, there were days where certain buses while they were already all 7 days of week, but more bus times were added, and therefore should not be counted in operating dates, instead will be calculated below in intervals
```

```{r}
intervalcalc = van_translink_trips %>%
  dplyr::select(trip_id, arrival_time, stop_id, stop_sequence, stop_name, trip_headsign, direction_id, service_id, stop_name)

intervalcalc$arrival_time <- as.numeric(hms(intervalcalc$arrival_time))

intervalcalc <- intervalcalc[order(intervalcalc$trip_headsign, intervalcalc$service_id, intervalcalc$stop_id, intervalcalc$arrival_time), ]

# Calculate the interval (difference in time) within each unique trip_headsign, service_id, stop_id trio
intervalcalc <- intervalcalc %>%
  group_by(trip_headsign, service_id, stop_id, stop_name) %>%
  mutate(interval = c(NA, diff(arrival_time))) %>%
  ungroup()

intervalcalc_filtered <- intervalcalc[!is.na(intervalcalc$interval), ]

# Calculate the average interval for each unique combination of trip_headsign, service_id, and stop_id
avg_intervals <- intervalcalc_filtered %>%
  group_by(trip_headsign, stop_id, stop_name) %>%
  summarise(avg_interval = mean(interval, na.rm = TRUE))
```

```{r}
van_translink_services <- van_translink_trips %>%
  # Convert arrival_time (hr:min:sec) to seconds
  mutate(arrival_time_sec = as.numeric(hms(arrival_time))) %>% 
  group_by(stop_id, direction_id, trip_headsign) %>%
  summarise(
    # Step 4: Include fields tied to stop_id
    stop_sequence = first(stop_sequence),
    stop_name = first(stop_name),
    stop_code = first(stop_code),
    stop_lat = first(stop_lat),
    stop_lon = first(stop_lon),
    service_id = first(service_id),
    block_id = first(block_id),
    shape_id = first(shape_id),
    route_id = first(route_id),
    .groups = "drop" # Ungroup after summarising
  ) 

van_translink_services = van_translink_services %>%
  left_join(unique_trip_sums, by = c("trip_headsign", "stop_id", "stop_name")) %>%
  left_join(avg_intervals, by = c("trip_headsign", "stop_id", "stop_name")) %>%
  mutate(pctOperating = operating_dates_sum/126*100)

van_translink_services = van_translink_services %>% #remove outliers
  filter(!is.na(avg_interval),
         avg_interval != 0.0)

van_translink_services = van_translink_services %>%
  mutate(frequency_score = pctOperating/100 * log(1+(max(van_translink_services$avg_interval)/avg_interval)))
```

```{r}
van_translink_station <- van_translink_services %>%
  group_by(stop_id, stop_name, stop_lat, stop_lon) %>%
  summarize(
    avg_interval = mean(avg_interval, na.rm = TRUE),
    pctOperating = mean(pctOperating, na.rm = TRUE),
    frequency_score = mean(frequency_score, na.rm = TRUE),
    .groups = "drop"  # Ungroup after summarization
  )
```

```{r}
van_indivbuffer = st_read("./data&code/200mIndivBuffer.geojson")

van_indivbuffer$stop_id = as.numeric(van_indivbuffer$stop_id)

van_databuffer = van_indivbuffer %>%
  left_join(van_translink_station, by = c("stop_id", "stop_name", "stop_lat", "stop_lon"))
```

```{r, eval=FALSE}
st_write(van_databuffer, "./data&code/van_databuffer.geojson")
```

```{r}
van_dataintersect = st_read("./data&code/van_dataintersect.geojson")

van_datacensus = van_dataintersect %>%
  group_by(CD_UID, name) %>%
  summarize(
    avg_interval = mean(avg_interval, na.rm = TRUE),
    pctOperating = mean(pctOperating, na.rm = TRUE),
    frequency_score = mean(frequency_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as.data.frame() %>%
  dplyr::select(-geometry)

van_datacensus = left_join(vancensus200, van_datacensus, by = c("CD_UID", "name"))
```

```{r, eval=FALSE}
st_write(van_datacensus, "./data&code/van_datacensus.geojson")
```

```{r fig.width = 10, fig.height=7}
ggplot(van_datacensus) +
  geom_sf(aes(fill = avg_interval)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())
```

```{r fig.width = 10, fig.height=7}
ggplot(van_datacensus) +
  geom_sf(aes(fill = pctOperating)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())
```

```{r fig.width = 10, fig.height=7}
ggplot(van_datacensus) +
  geom_sf(aes(fill = frequency_score)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())
```

```{r}
van_datacensus = van_datacensus %>%
  mutate(minority_z = (minorityPct - mean(van_datacensus$minorityPct))/sd(van_datacensus$minorityPct),
         disadvantaged_z = (disadvPopPct - mean(van_datacensus$disadvPopPct))/sd(van_datacensus$disadvPopPct),
         medinc_z = (medinc - mean(van_datacensus$medinc))/sd(van_datacensus$medinc) * -1,
         marginalized_score = minority_z + disadvantaged_z + medinc_z)
```

```{r fig.width = 10, fig.height=7}
finaldata <- bi_class(van_datacensus, x = frequency_score, y = marginalized_score, style = "quantile", dim = 3)

finalmap <- ggplot() +
  geom_sf(data = finaldata, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "GrPink", dim = 3) +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) +
  bi_theme()

finallegend = bi_legend(pal = "GrPink",
                    dim = 3,
                    xlab = "Higher Frequency",
                    ylab = "Higher Marginalized",
                    size = 6)

finalplot <- ggdraw() +
  draw_plot(finalmap, 0, 0, 1, 1) +
  draw_plot(finallegend, 0, 0.15, 0.2, 0.2)

finalplot
```

```{r}
lm_model <- lm(frequency_score ~ marginalized_score, data = van_datacensus)

summary(lm_model)

plot(van_datacensus$marginalized_score, van_datacensus$frequency_score, 
     main = "Vancouver", 
     xlab = "Underprivileged Score", 
     ylab = "Frequency Score")

abline(lm_model, col = "red")
```
