---
title: "TransportationFinal"
author: "Jack Chen"
date: "2024-12-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidycensus)
library(dplyr)
library(cancensus)
library(ggplot2)
library(RColorBrewer)
library(tidyr)
library(knitr)
library(tidyverse)
library(kableExtra)
library(terra)
library(sf)
library(biscale)
library(cowplot)
```
Vector: short variable code
Type: variables are provided as aggregates of female responses, male responses, or total (male+female) responses
Label: detailed variable name
Units: provides information about whether the variable represents a count integer, a ratio, a percentage, or a currency figure
Parent_vector: shows the immediate hierarchical parent category for that variable, where appropriate
Aggregation: indicates how the variable should be aggregated with others, whether it is additive or if it is an average of another variable
Description: a rough description of a variable based on its hierarchical structure. This is constructed by cancensus by recursively traversing the labels for every variable’s hierarchy, and facilitates searching for specific variables using key terms.     

```{r}
acs_varlist = c("B02001_001","B02001_003",# total, black
                        "B02001_005", # asian,
                        "B02001_004", "B02001_006", "B02001_007", # first nation, non asian aapi, other races
                        "B19013_001",  # hhinc
                        "B01001_003", "B01001_027", # <5 male/female, 
                        "B01001_004", "B01001_028", # 5-9 male/female, 
                        "B01001_005", "B01001_029", # 10-14 male/female
                        "B01001_020", "B01001_044", # 65+ male/female v
                        "B01001_021", "B01001_045",
                        "B01001_022", "B01001_046",
                        "B01001_023", "B01001_047",
                        "B01001_024", "B01001_048",
                        "B01001_025", "B01001_049")
```

```{r}
cooktracts =  
  get_acs(geography = "tract",
          variables = acs_varlist, 
          year=2021, state="IL",
          county="Cook",
          geometry=FALSE) 
```

```{r}
chitracts = read.csv("./data&code/chicagotracts.csv") %>%
  rename(geometry = the_geom) %>%
  st_as_sf(wkt = "geometry", crs = 4326)
```

```{r}
cooktracts = cooktracts %>%
  dplyr::select(-NAME, -moe) %>%
  spread(key = variable, value = estimate) %>%
  rename(totPop21 = B02001_001,
          totBlack = B02001_003,
          totAsian = B02001_005,
          totFirstNations = B02001_004,
          totAAPI = B02001_006,
          totOther = B02001_007,
          medinc = B19013_001,
          male5 = B01001_003,
          fem5 = B01001_027, # <5 male/female, 
          male9 = B01001_004,
          fem9 = B01001_028, # 5-9 male/female, 
          male14 = B01001_005,
          fem14 = B01001_029, # 10-14 male/female
          male66 = B01001_020,
          fem66 = B01001_044, # 65+ male/female v
          male69 = B01001_021,
          fem69 = B01001_045,
          male74 = B01001_022,
          fem74 = B01001_046,
          male79 = B01001_023, 
          fem79 = B01001_047,
          male84 = B01001_024, 
          fem84 = B01001_048,
          male85 = B01001_025,
          fem85 = B01001_049) %>%
mutate(totMinority = totBlack + totAsian + totFirstNations + totAAPI + totOther,
       minorityPct = totMinority/totPop21,
       disadvPop = male5+fem5+male9+fem9+male14+fem14+male66+fem66+male69+fem69+male74+fem74+male79+fem79+male84+fem84+male5+fem85,
       disadvPopPct = disadvPop/totPop21) 

cooktracts = cooktracts %>%
dplyr::select(-totBlack,-totAsian,-totFirstNations,-totAAPI,-totOther,-male5,-fem5,-male9,-fem9,-male14,-fem14,-male66,-fem66,-male69,-fem69,-male74,-fem74,-male79,-fem79,-male84,-fem84,-male5,-fem85)
```

```{r}
ggplot() + 
  geom_sf(data = chitracts) +
  geom_sf(data = vanstations) +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) 
```


```{r fig.width = 10, fig.height = 8}
cooktracts$GEOID = as.numeric(cooktracts$GEOID)

chitracts.f = chitracts %>%
  rename(GEOID = GEOID10) %>%
  left_join(as.data.frame(cooktracts), by = "GEOID") %>%
  na.omit()
```

```{r eval=FALSE}
st_write(chitracts.f, "./data&code/chitracts_f.geojson", append=FALSE)
```

```{r}
ctastops = read.csv("./data&code/ctadata/stops.csv")
chistops = st_as_sf(ctastops, coords = c("stop_lon", "stop_lat"), crs = 4326)
```

```{r}
st_write(chistops, "./data&code/chistops.geojson")
```

```{r fig.width = 8, fig.height = 10}
chistations = st_read("./data&code/chistations.geojson")

ggplot() + 
  geom_sf(data = chitracts.f) +
  geom_sf(data = chistations) +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) 
```

```{r}
intersection200 = st_read("./data&code/chiIntersect200.geojson")
head(as.data.frame(intersection200))

chitracts200 = chitracts.f %>%
  left_join(as.data.frame(intersection200) %>%
              select(NAME10, percentagecover, GEOID, area, intersectarea, -NOTES), by = "GEOID") %>%
  filter(!is.na(percentagecover),
         !is.na(medinc),
         !is.na(totPop21)) %>%
  na.omit() %>%
  mutate("covPop" = intersectarea/totPop21,
         "covPctPop" = percentagecover/totPop21) %>%
  mutate("weightedCover" = percentagecover*(totPop21/(sum(chitracts200$totPop21))))
```

```{r eval=FALSE}
st_write(chitracts200, "./data&code/chitracts.geojson", append=FALSE)
```

```{r}
ggplot(chitracts200) +
  geom_sf(aes(fill = weightedCover)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())
```

```{r fig.width = 10, fig.height=7}
minoritycover <- bi_class(chitracts200, x = weightedCover, y = minorityPct, style = "quantile", dim = 3)

minoritymap <- ggplot() +
  geom_sf(data = minoritycover, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "GrPink", dim = 3) +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) +
  bi_theme()

minoritylegend = bi_legend(pal = "GrPink",
                    dim = 3,
                    xlab = "Higher % Station Cover ",
                    ylab = "Higher % Minority ",
                    size = 6)

minorityPlot <- ggdraw() +
  draw_plot(minoritymap, 0, 0, 1, 1) +
  draw_plot(minoritylegend, 0, 0.15, 0.2, 0.2)

minorityPlot
```

```{r fig.width = 10, fig.height=7}
disadvcover <- bi_class(chitracts200, x = weightedCover, y = disadvPopPct, style = "quantile", dim = 3)

disadvmap <- ggplot() +
  geom_sf(data = disadvcover, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "GrPink", dim = 3) +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) +
  bi_theme()

disadvlegend = bi_legend(pal = "GrPink",
                    dim = 3,
                    xlab = "Higher % Station Cover",
                    ylab = "Higher % Disadvantaged",
                    size = 6)

disadvPlot <- ggdraw() +
  draw_plot(disadvmap, 0, 0, 1, 1) +
  draw_plot(disadvlegend, 0, 0.15, 0.2, 0.2)

disadvPlot
```

```{r fig.width = 10, fig.height=7}
inccover <- bi_class(chitracts200, x = weightedCover, y = medinc, style = "quantile", dim = 3)

incmap <- ggplot() +
  geom_sf(data = inccover, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "GrPink", dim = 3) +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) +
  bi_theme()

inclegend = bi_legend(pal = "GrPink",
                    dim = 3,
                    xlab = "Higher % Station Cover",
                    ylab = "Higher Median Income",
                    size = 6)

incPlot <- ggdraw() +
  draw_plot(incmap, 0, 0, 1, 1) +
  draw_plot(inclegend, 0, 0.15, 0.2, 0.2)

incPlot
```

```{r}
routes = read.csv("./data&code/ctadata/routes.csv")

trips = read.csv("./data&code/ctadata/trips.csv")

bus_time = trips %>%
  left_join(routes, by = "route_id") %>%
  dplyr::select(-route_url, -route_color, -route_text_color)
```

```{r}
stop_times = read.csv("./data&code/ctadata/stop_times.csv")

chistations$stop_id = as.numeric(chistations$stop_id)

head(as.data.frame(chistations))

chi_stop_times <- stop_times %>%
  left_join(chistations, by = "stop_id") %>%
  dplyr::select(-stop_code, -parent_station, -pickup_type) 
```

```{r}
trips_filtered <- bus_time %>%
  semi_join(chi_stop_times, by = "trip_id")  # Keeps only trip_ids in stop_times_with_info

chi_stop_trips <- chi_stop_times %>%
  left_join(trips_filtered %>% select(route_id, service_id, trip_id, route_short_name, route_long_name, direction_id, block_id, shape_id),
            by = "trip_id") %>%
  filter(!is.na(stop_name))

chi_stop_trips = chi_stop_trips %>%
  dplyr::select(-shape_dist_traveled, -location_type, -stop_desc)
```

```{r}
calendar = read.csv("./data&code/ctadata/calendar.csv")

chi_cal <- chi_stop_trips %>%
  left_join(calendar %>% select(service_id, monday, tuesday, wednesday, thursday, friday, saturday, sunday),
            by = "service_id") %>%
  mutate(weekday_avail = monday+tuesday+wednesday+thursday+friday+saturday+sunday)
```

```{r}
dates = read.csv("./data&code/ctadata/calendar_dates.csv")

# Step 1: Create a count of exceptions for service_id in the dates dataframe
exception_counts <- dates %>%
  filter(exception_type == 1) %>%
  group_by(service_id) %>%
  summarize(total_exceptions = n(), .groups = "drop")  

# Step 2: Add operating_dates to final_df_with_calendar
chi_cta_trips <- chi_cal %>%
  # Left join with exception counts
  left_join(exception_counts, by = "service_id") %>%
  # Create the operating_dates column based on conditions
  mutate(
    total_exceptions = case_when(!is.na(total_exceptions) ~ total_exceptions,
                                 TRUE ~ 0),
    operating_dates = case_when(monday == 1 & tuesday == 1 ~ total_exceptions + 2 + weekday_avail*11,
                                monday == 1 & tuesday == 0 ~ total_exceptions + 1 + weekday_avail*11,
                                monday == 0 & tuesday == 1 ~ total_exceptions + 1 + weekday_avail*11,
                                TRUE ~ total_exceptions + weekday_avail*11),
    pctOperating = operating_dates/79*100
  ) %>%
  # Drop the total_exceptions column (optional, if not needed further)
  select(-total_exceptions)
```

```{r}
unique_trips = chi_cta_trips %>%
  dplyr::select(-arrival_time, -trip_id, -departure_time, -block_id, -monday, -tuesday, -wednesday, -thursday, -friday, -saturday, -sunday, -shape_id, -stop_sequence) %>%
  unique() #%>%
  #dplyr::select(-service_id)

unique_trip_sums <- unique_trips %>%
  group_by(route_long_name, stop_id, stop_name) %>%
  summarise(
    weekday_avail_sum = sum(weekday_avail, na.rm = TRUE),
    operating_dates_sum = sum(operating_dates, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(operating_dates_sum = case_when(operating_dates_sum > 79 ~ 79,
                                        TRUE ~ operating_dates_sum)) 
# Upon inspection, there were days where certain buses while they were already all 7 days of week, but more bus times were added, and therefore should not be counted in operating dates, instead will be calculated below in intervals
```

```{r}
intervalcalc = chi_cta_trips %>%
  dplyr::select(trip_id, arrival_time, stop_id, stop_sequence, stop_name, route_long_name, direction_id, service_id, stop_name)

intervalcalc$arrival_time <- as.numeric(hms(intervalcalc$arrival_time))

intervalcalc <- intervalcalc[order(intervalcalc$route_long_name, intervalcalc$service_id, intervalcalc$stop_id, intervalcalc$arrival_time), ]

# Calculate the interval (difference in time) within each unique route_long_name, service_id, stop_id trio
intervalcalc <- intervalcalc %>%
  group_by(route_long_name, service_id, stop_id, stop_name) %>%
  mutate(interval = c(NA, diff(arrival_time))) %>%
  ungroup()

intervalcalc_filtered <- intervalcalc[!is.na(intervalcalc$interval), ]

# Calculate the average interval for each unique combination of route_long_name, service_id, and stop_id
avg_intervals <- intervalcalc_filtered %>%
  group_by(route_long_name, stop_id, stop_name) %>%
  summarise(avg_interval = mean(interval, na.rm = TRUE))
```

```{r}
chi_cta_services <- chi_cta_trips %>%
  # Convert arrival_time (hr:min:sec) to seconds
  mutate(arrival_time_sec = as.numeric(hms(arrival_time))) %>% 
  group_by(stop_id, direction_id, route_long_name) %>%
  summarise(
    # Step 4: Include fields tied to stop_id
    stop_sequence = first(stop_sequence),
    stop_name = first(stop_name),
    stop_headsign = first(stop_headsign),
    service_id = first(service_id),
    block_id = first(block_id),
    shape_id = first(shape_id),
    route_id = first(route_id),
    .groups = "drop" # Ungroup after summarising
  ) 

chi_cta_services = chi_cta_services %>%
  left_join(unique_trip_sums, by = c("route_long_name", "stop_id", "stop_name")) %>%
  left_join(avg_intervals, by = c("route_long_name", "stop_id", "stop_name")) %>%
  mutate(pctOperating = operating_dates_sum/79*100)

chi_cta_services = chi_cta_services %>% #remove outliers
  filter(!is.na(avg_interval),
         avg_interval != 0.0)

chi_cta_services = chi_cta_services %>%
  mutate(frequency_score = pctOperating/100 * log(1+(max(chi_cta_services$avg_interval)/avg_interval)))
```

```{r}
chi_cta_station <- chi_cta_services %>%
  group_by(stop_id, stop_name) %>%
  summarize(
    avg_interval = mean(avg_interval, na.rm = TRUE),
    pctOperating = mean(pctOperating, na.rm = TRUE),
    frequency_score = mean(frequency_score, na.rm = TRUE),
    .groups = "drop"  # Ungroup after summarization
  )
```

```{r}
chi_indivbuffer = st_read("./data&code/Chi200IndivBuff.geojson")

chi_indivbuffer$stop_id = as.numeric(chi_indivbuffer$stop_id)

chi_databuffer = chi_indivbuffer %>%
  left_join(chi_cta_station, by = c("stop_id", "stop_name"))
```

```{r, eval=FALSE}
st_write(chi_databuffer, "./data&code/chi_databuffer.geojson", append=FALSE)
```

```{r}
chi_dataintersect = st_read("./data&code/chi_dataintersect.geojson")

chi_datacensus = chi_dataintersect %>%
  as.data.frame() %>%
  dplyr::select(-geometry) %>%
  group_by(GEOID, NAME10) %>%
  summarize(
    avg_interval = mean(avg_interval, na.rm = TRUE),
    pctOperating = mean(pctOperating, na.rm = TRUE),
    frequency_score = mean(frequency_score, na.rm = TRUE),
    .groups = "drop"
  ) 

chi_datacensus = left_join(chitracts200 %>% rename(NAME10 = NAME10.x), chi_datacensus, by = c("GEOID", "NAME10"))
```

```{r, eval=FALSE}
st_write(chi_datacensus, "./data&code/chi_datacensus.geojson")
```

```{r fig.width = 10, fig.height=7}
ggplot(chi_datacensus) +
  geom_sf(aes(fill = avg_interval)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())
```

```{r fig.width = 10, fig.height=7}
ggplot(chi_datacensus) +
  geom_sf(aes(fill = pctOperating)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())
```

```{r fig.width = 10, fig.height=7}
ggplot(chi_datacensus) +
  geom_sf(aes(fill = frequency_score)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())
```

```{r}
chi_datacensus = chi_datacensus %>%
  mutate(minority_z = (minorityPct - mean(chi_datacensus$minorityPct))/sd(chi_datacensus$minorityPct),
         disadvantaged_z = (disadvPopPct - mean(chi_datacensus$disadvPopPct))/sd(chi_datacensus$disadvPopPct),
         medinc_z = (medinc - mean(chi_datacensus$medinc))/sd(chi_datacensus$medinc) * -1,
         marginalized_score = minority_z + disadvantaged_z + medinc_z)
```

```{r fig.width = 10, fig.height=7}
finaldata <- bi_class(chi_datacensus, x = frequency_score, y = marginalized_score, style = "quantile", dim = 3)

finalmap <- ggplot() +
  geom_sf(data = finaldata, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "GrPink", dim = 3) +
  theme(axis.text = element_blank(),    
        axis.ticks = element_blank(),   
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank()) +
  bi_theme()

finallegend = bi_legend(pal = "GrPink",
                    dim = 3,
                    xlab = "Higher Frequency",
                    ylab = "Higher Marginalized",
                    size = 6)

finalplot <- ggdraw() +
  draw_plot(finalmap, 0, 0, 1, 1) +
  draw_plot(finallegend, 0, 0.15, 0.2, 0.2)

finalplot
```

```{r}
lm_model <- lm(frequency_score ~ marginalized_score, data = chi_datacensus)

summary(lm_model)

plot(chi_datacensus$marginalized_score, chi_datacensus$frequency_score, 
     main = "Chicago", 
     xlab = "Underprivileged Score", 
     ylab = "Frequency Score")

abline(lm_model, col = "red")
```

